---
title: "Explore Citi Bike Data"
output:
  html_notebook: default
  pdf_document: default
  html_document:
    df_print: paged
---

### Citi Bike Data

The data is from Citi Bike website containing information about individual trips from January of 2020.

More data on https://www.citibikenyc.com/system-data

The data in use is a small subset. 


```{r}
# Load the data set into a data frame

all_data <- read.csv("january_trips_subset.csv")
```


```{r}
# Investigate the data

head(all_data)
colnames(all_data)
```


### Citi Bike Data
Investing data by generating a heat map.

```{r}
# Install and load ggplot2 and dplyr
library(ggplot2)
library(dplyr)
# Create a heatmap
station_heat_map <- ggplot(all_data, aes(x=start.station.longitude,y=start.station.latitude))+
  geom_bin2d(binwidth= c(0.003,0.003))

station_heat_map

```


### Modifying the Data Frame: Adding Age

Adding a column called `age` to the data frame. `age` 

```{r}
# Add the age column
all_data<- all_data %>% 
  mutate(age= 2020-birth.year)
head(all_data)


```

### Calculate Distance

Using Google Map Api to caclulate biked distance. Test as below, api key is hidden. 

```{r}
# Try creating a distance column in your data frame here:

library(ggmap)
library(gmapsdistance)

all_data<- all_data %>% 
  mutate(start= paste(start.station.latitude+ ","+ start.station.longitude))%>%
  mutate(end= paste(end.station.latitude+ ","+ end.station.longitude))%>%


set.api.key("$$$$$$")
results = gmapsdistance(origin = all_data.start, 
                        destination = all_data.end, 
                       mode = "bicycling", key = get.api.key())

results

```

Below is a simplified version of just finding the shortest distance between two geo locations without using Google Map Api.

```{r}
# Use the geosphere library to create a distance column
library(geosphere)
library(dplyr)
starting_stations <- all_data %>% select(start.station.longitude, start.station.latitude)
ending_stations <- all_data %>% select(end.station.longitude, end.station.latitude)
all_data <- all_data%>% mutate(distance=distHaversine(starting_stations,ending_stations))

```

### Cleaning up and adding Speed

Clean up the data and adding speed column

```{r}
# Create the speed column

all_data <- all_data%>% filter(age<80)%>%  filter(age>10)%>%mutate(speed= distance/tripduration)
all_data
```

### Average Speed by Age



```{r}
# Use group_by() and summarize() to get the mean speed of each age
average_speed_by_age <- all_data %>% 
  group_by(age)%>%
  summarise(mean_speed= mean(speed))

average_speed_by_age

```

### Visualization of age and speed relationship

```{r}
# Install and load ggplot2 to create a line graph of age and mean speed
speed_by_age <- ggplot(data=average_speed_by_age, aes(x=age, y=mean_speed))+
  geom_line()+
  scale_x_continuous(breaks = seq(0,100,by=10))+ 
  labs(title = "Average speed of Citi Bike users by age (January 2020)", x = "Age", y = "Average Speed (m/s)") + 
  theme(plot.title = element_text(hjust = 0.5))
speed_by_age
```


### Filtering By Gender


```{r}
# Use group_by() again to group by both age and gender
average_speed_by_age_and_gender<- all_data %>% 
  group_by(age, gender)%>%
  summarise(mean_speed= mean(speed))
average_speed_by_age_and_gender
```

Visulize data:

```{r}
# Make a line graph of filtered data frame
speed_by_age_gender <- ggplot(data=average_speed_by_age_and_gender, aes(x=age, y=mean_speed, color=gender))+
  geom_line()+
  scale_x_continuous(breaks = seq(0,100,by=10))+ 
  labs(title = "Average speed of Citi Bike users by age and gender (January 2020)", x = "Age", y = "Average Speed (m/s)") + 
  theme(plot.title = element_text(hjust = 0.5))

# Use mutate() and as.factor() to change the gender column into a factor.
speed_by_age_gender <- ggplot(data=average_speed_by_age_and_gender, aes(x=age, y=mean_speed, color=as.factor(gender)))+
  geom_line()+
  scale_x_continuous(breaks = seq(0,100,by=10))+ 
  labs(title = "Average speed of Citi Bike users by age and gender (January 2020)", x = "Age", y = "Average Speed (m/s)", color='Gender') + 
  theme(plot.title = element_text(hjust = 0.5))
speed_by_age_gender

```




###Clean up the data to illustrate relationship between speed and gender

```{r}

# Filter the data frame to only include genders 1 and 2. Set appropriate labels for the legend

average_speed_by_age_and_gender %>% filter(age < 80, gender == 1 | gender == 2) %>% ggplot() + geom_line(aes(x = age, y = mean_speed, color=as.factor(gender))) + labs(title = "Average speed of Citi Bike users by age (January 2020)", x = "Age", y = "Average Speed (m/s)") + theme(plot.title = element_text(hjust = 0.5)) + scale_color_discrete(name = "Gender", labels = c("Male Identifying", "Female Identifying"))
```



